---
title: "Covid '19 Analysis"
output:
  html_document: default
  pdf_document: default
date: "2025-01-29"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Fellow CU Boulder peers reviewing these work, scroll down to Costa Rica COVID19 Data Analysis to skip the code we already reviewed in class


```{r import libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
```



```{r read data}

url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

global_cases <- url %>% 
  str_c("time_series_covid19_confirmed_global.csv") %>% 
  read_csv()

global_deaths <- url %>% 
  str_c("time_series_covid19_deaths_global.csv") %>% 
  read_csv()

us_cases <- url %>% 
  str_c("time_series_covid19_confirmed_US.csv") %>% 
  read_csv()

us_deaths <- url %>% 
  str_c("time_series_covid19_deaths_US.csv") %>% 
  read_csv()

```

```{r tidy data}
cols <- c('Province/State', 'Country/Region', 'Lat', 'Long')

global_cases <- global_cases %>% 
    pivot_longer(cols = -cols,
                 names_to = "date",
                 values_to = "cases") %>% 
    select(-c(Lat,Long)) %>% 
  clean_names()

global_deaths <- global_deaths %>% 
    pivot_longer(cols = -cols,
                 names_to = "date",
                 values_to = "deaths") %>% 
    select(-c(Lat,Long)) %>% 
  clean_names()

```

```{r Transforming Global}
global <- global_cases %>% 
  full_join(global_deaths) %>% 
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

summary(global)

global %>% filter(cases > 28000000, country_region == "US")

```

```{r Transforming US Cases}
us_cases <- us_cases %>% 
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>% 
  select(Admin2:cases) %>% 
  mutate(date = mdy(date)) %>% 
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>% 
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>% 
  select(Admin2:deaths) %>% 
  mutate(date = mdy(date)) %>% 
  select(-c(Lat, Long_))

us <- us_cases %>% 
  full_join(us_deaths) %>% 
  clean_names()
us

```

```{r Adding population to global}
global <- global %>% 
  unite("combined_key",
        c(province_state, country_region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)


uid <- str_c(
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/",
        "UID_ISO_FIPS_LookUp_Table.csv") %>% 
  read_csv() %>% 
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2)) %>% 
  clean_names()

global <- global %>% 
  left_join(uid, by = c("province_state", "country_region")) %>% 
  select(province_state, country_region, date, cases, deaths, 
         population, combined_key)

```

```{r Visualizing Data}
us_by_state <- us %>% 
  group_by(province_state, country_region, date) %>% 
  summarise(cases = sum(cases),
            deaths = sum(deaths),
            population = sum(population)) %>% 
  mutate(deaths_per_mill = deaths*1000000/population) %>% 
  ungroup()
us_by_state

us_totals <- us_by_state %>% 
  group_by(country_region, date) %>% 
  summarise(cases = sum(cases),
            deaths = sum(deaths),
            population = sum(population)) %>% 
  mutate(deaths_per_mill = deaths*1000000/population) %>% 
  ungroup()

us_totals %>% 
  filter(cases > 0) %>% 
  ggplot(aes(date, cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color="deaths")) +
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in US", y=NULL)

us_by_state %>% 
  filter(province_state == "New York") %>% 
  filter(cases > 0) %>% 
  ggplot(aes(date, cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color="deaths")) +
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in New York", y=NULL)

```

```{r Analyze Data}
us_by_state <- us_by_state %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
us_totals <- us_totals %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_totals %>% 
  ggplot(aes(date, new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color="new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in US", y=NULL)

us_state_totals <- us_by_state %>% 
  group_by(province_state) %>% 
  summarize(deaths = max(deaths),
            cases = max(cases),
            population = max(population),
            cases_per_thou = 1000* cases/population,
            deaths_per_thou = 1000* deaths/population) %>% 
  filter(cases>0, population>0)
  
us_state_totals %>% 
  slice_min(deaths_per_thou, n=10)

us_state_totals %>% 
  slice_max(deaths_per_thou, n=10)

```

```{r Modeling Data}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

us_tot_w_pred <- us_state_totals %>% 
  mutate(pred = predict(mod))

us_tot_w_pred %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

# Costa Rica COVID19 Data Analysis

For this analysis, you will learn more about the COVID19 Pandemic as it develop in Costa Rica.

```{r Transforming CR}
cr <- global %>% 
  filter(country_region == "Costa Rica") %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths)) %>% 
  select(-c(province_state, country_region, combined_key)) # Dropping province_state because all NAs

cr_monthly <- cr %>% 
  group_by(month = floor_date(date, unit = "month")) %>% 
  summarize(cases = sum(cases),
            new_cases = sum(new_cases),
            deaths = sum(deaths),
            new_deaths = sum(new_deaths),
            population = max(population)) %>% 
  ungroup()


cr_monthly
```

## Visualizing Costa Rica during the pandemic

```{r Visualizing 1}
cr_monthly %>% 
  ggplot(aes(x = month)) +
  geom_line(aes(y = new_cases, color = "New Cases"), size = 1.2) +  # Line for new cases
  geom_line(aes(y = new_deaths, color = "Deaths"), size = 1.2) +        # Line for deaths
  scale_y_log10() +  # Log10 scale for y-axis with comma formatting
  labs(
    title = "Costa Rica monthly COVID-19 new cases and new deaths",
    subtitle = "Tracking the rise and fall of new cases and new deaths over time",
    x = "Month",
    y = "Count (Log10 Scale)",
    color = "Metric",  # Legend title
    caption = "Data Source: Your Data Source Here"
  ) +
  scale_color_manual(values = c("New Cases" = "darkgreen", "Deaths" = "red")) +  # Custom colors
  theme_minimal() +  # Start with a minimal theme
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "black", hjust = 0.5),
    plot.subtitle = element_text(size = 14, color = "gray40", hjust = 0.5),
    plot.caption = element_text(size = 10, color = "gray40", hjust = 1),
    axis.title = element_text(size = 14, face = "bold", color = "black"),
    axis.text = element_text(size = 12, color = "black"),
    panel.background = element_rect(fill = "white"),  # Light background
    plot.background = element_rect(fill = "white"),   # Light background
    panel.grid.major = element_line(color = "gray90"),  # Light grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    text = element_text(color = "black"),  # Black text for contrast
    legend.position = "bottom"  # Move legend to the bottom
  )+
  geom_smooth(aes(y = new_cases, color = "New Cases Trend"), 
              method = "lm", formula = y ~ poly(x, 2), se = FALSE, linetype = "dashed") +  # Quadratic trend for cases
  geom_smooth(aes(y = new_deaths, color = "Deaths Trend"), 
              method = "lm", formula = y ~ poly(x, 2), se = FALSE, linetype = "dashed")   # Quadratic trend for deaths
```

The quadratic trend for the deaths seems to have a stronger curvature indicating the deaths dropped faster than the new cases and will be explored further with in the modeling part.

But first, a break down analysis of the seasonal trend, to identify if their were specific months higher than others.

```{r Visualizing 2}
cr_monthly %>%
  mutate(
    month_num = as.factor(month(month, label = TRUE, abbr = TRUE)),  # Convert month to ordered factor
    year = as.factor(year(month))  # Convert year to factor
  ) %>%
  ggplot(aes(x = month_num, y = new_deaths, color = year, group = year)) +  # Group by year
  geom_line(size = 1.2) +  # Add lines
  labs(
    title = "Costa Rica monthly COVID-19 deaths by year",
    subtitle = "Trends in new deaths over time, grouped by year",
    x = "Month",
    y = "New Deaths",
    color = "Year",  # Legend title
  ) +
  theme_minimal() +  # Start with a minimal theme
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "black", hjust = 0.5),
    plot.subtitle = element_text(size = 14, color = "gray40", hjust = 0.5),
    plot.caption = element_text(size = 10, color = "gray40", hjust = 1),
    axis.title = element_text(size = 14, face = "bold", color = "black"),
    axis.text = element_text(size = 12, color = "black"),
    panel.background = element_rect(fill = "white"),  # Light background
    plot.background = element_rect(fill = "white"),   # Light background
    panel.grid.major = element_line(color = "gray90"),  # Light grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    text = element_text(color = "black"),  # Black text for contrast
    legend.position = "bottom"  # Move legend to the bottom
  )
```

The graph shows May and September of 2021 as the deadliest months of the COVID19th pandemic in Costa Rica.

## Modeling Costa Rica

### Simple linear model

```{r}
cr_monthly <- cr_monthly %>%
  mutate(days_since_start = as.numeric(month - min(month)))
mod_lin <- lm(new_deaths ~ days_since_start, cr_monthly)
summary(mod_lin)
```

A simple linear model of the new_deaths over time shows a weak fit, the key takeaways are:

- The model explains only 8.5% of the variance in deaths (R-squared=0.085).

- The negative coefficient for days_since_start (-0.21) suggests a slight decrease in deaths over time, but this effect is not statistically significant (p=0.084).

Time alone, as a linear predictor, does not meaningfully explain trends in COVID-19 deaths.

### Quadratic Model

```{r}
mod_quad <- lm(new_deaths ~ days_since_start + I(days_since_start^2), cr_monthly)
summary(mod_quad)
```

By including the time variable as quadratic predictor, a strong improvement is found. They key differences are:

- The model explains 42.7% of the variance (R-squared=0.427), a 5x improvement over the linear model.

- Both the linear term (p=0.0007) and quadratic term (p<0.0001) are highly significant.

Interpretation:

- Positive linear coefficient (1.49): Deaths initially increase with time.

- Negative quadratic coefficient (-0.0015): After a peak, deaths decrease over time.

- The model captures a wave-like pattern (e.g., a surge followed by decline), common in pandemics like COVID-19.


## Conclusion

After modeling the data, a non-Linear trend was discovered. The quadratic term helped in capturing the true relationship between time and deaths. Further analysis and more data is needed to arrive at a strong conclusion but the wave-like pattern could reflect public health interventions (e.g., lock downs, vaccinations) impacting death rates over time.