---
title: "Shooting Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2025-01-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(janitor)
```

# Importing the data

```{r Importing the data}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
shootings_raw <- read_csv(url) %>% 
  clean_names() # Janitor turns every column name to snake_case
```

# Describing the data

```{r Describing the data}
shootings_raw %>% glimpse()
```

### Key variables definition:

-   boro: Borough where the shooting incident occurred.
-   precinct: a district of a city or town as defined for police purposes.
-   jurisdiction code: Jurisdiction where the shooting incident occurred. Jurisdiction codes 0(Patrol), 1(Transit) and 2(Housing) represent NYPD whilst codes 3 and more represent non NYPD jurisdictions.
-   statistical murder flag: Shooting resulted in the victim’s death which would be counted as a murder.

Note: Other variables were considered self-explanatory according to their names.

# Transforming the data

```{r Transforming the data}
shootings_raw %>% count(jurisdiction_code)
shootings_raw %>% count(loc_classfctn_desc)
shootings_raw %>% count(location_desc, sort = TRUE)

shootings <- shootings_raw %>% 
  select(-c(loc_of_occur_desc,
            loc_classfctn_desc,
            lon_lat)) %>% 
  mutate(jurisdiction_code = as_factor(jurisdiction_code),
         jurisdiction_code = fct_recode(
           jurisdiction_code, 
           "Patrol" = "0", 
           "Transit" = "1",
           "Housing" = "2"),
         occur_date = mdy(occur_date),
         precinct = as_factor(precinct))

shootings %>% count(jurisdiction_code)
shootings %>% glimpse()
```

-   Dropping *loc_of_occur_desc*, *loc_classfctn_desc* because too many values were missing. Also dropped *lon_lat* because they're available separately.

-   Re-coded *jurisdiction_code* based on description values.

-   Corrected *occur_date* format.

-   Converted precinct into factor to prevent treating it as a numeric variable.

# Visualizing the data

```{r Visualizing the Data 1}
defined_sex = c("M", "F")

shootings %>% 
  filter(perp_sex %in% defined_sex,
         vic_sex %in% defined_sex) %>% 
  count(perp_sex, vic_sex) %>% 
  ggplot(aes(x = perp_sex, y = n, fill = vic_sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +  
  labs(
    title = "Shootings by Perpetrator Sex and Victim Sex",
    x = "Perpetrator Sex",
    y = "Number of Shootings",
    fill = "Victim Sex"
  ) +
  theme_minimal() +  # Clean and minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16), 
    axis.title = element_text(size = 14),  # Axis titles
    axis.text = element_text(size = 12),   # Axis labels
    legend.title = element_text(size = 14),  # Legend title
    legend.text = element_text(size = 12)   # Legend labels
  )
```

The first plot shows the dominance of male over female perpetrators, while still not clear if there is an effect of male perpetrators against female victims.

```{r Visualizing the Data 2}
shootings %>% 
  filter(perp_sex %in% defined_sex,
         vic_sex %in% defined_sex,
        ) %>% 
  mutate(month = floor_date(occur_date, "month")) %>%
  count(perp_sex, vic_sex, month) %>% 
  ggplot(aes(x = month, y = n, color = interaction(perp_sex, vic_sex))) +
  geom_line(alpha=0.5) +  # Use geom_line() for a line chart
  scale_y_log10()+
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +  # Add linear trend lines
  labs(
    title = "Monthly Shootings by Perpetrator vs Victim sex Over Time",
    x = "Month",
    y = "Total Number of Shootings log(n)",
    color = "Perp-Vic Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

In the second plot, you can observe the decrease of shootings over time overall, while female against female shootings have the only positive trend of all four comparisons.

```{r Visualizing the Data 3}
shootings %>% 
  filter(vic_sex == "F") %>% 
  ggplot() +
  geom_point(aes(x = longitude, y = latitude, color = statistical_murder_flag), size = 3, alpha = 0.8) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "lightgray")) +
  labs(
    title = "Female Shooting Victims by location in New York City",
    x = "Longitude",
    y = "Latitude",
    color = "Is Murder?"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

```

Finally, you can look at a map of the shootings filtered by Female victims. Where there is no clear safe zone of less murders occuring due to the red dots being distributed across all the map.

# Modeling the data

A model is provided below to further analyze the effect of sex over the number of shootings over time.

```{r}
mod_data <- shootings %>% 
  filter(perp_sex %in% defined_sex,
         vic_sex %in% defined_sex,
        ) %>% 
  mutate(month = floor_date(occur_date, "month")) %>%
  mutate(month_num = as.numeric(month - min(month)) / 30.44) %>%   # Months since start
  count(perp_sex, vic_sex, month_num)
mod <- lm(data = mod_data, n ~ month_num + perp_sex * vic_sex)
summary(mod)
```

## Key Model takeways

Time Trend: Shootings decrease by \~0.087 per month (p\<0.001), suggesting a small but significant downward trend.

Sex Effects:

Male Perpetrators: Associated with significantly higher shootings when targeting female victims (baseline).

Male Victims: No significant effect when perpetrators are female (p=0.87).

Interaction: Male perpetrators targeting male victims drive a massive increase in shootings (58.10/month), indicating a strong synergistic effect.

Baseline: Female perpetrators targeting female victims are associated with \~11 shootings/month.

# Conclusion

With a quick glimpse of the data, you can identify males as a clear participant of New York City shootings. Since this could be a source of bias, models were interpreted to spotlight the effect of female perpetrators and victims. There was not a clear effect of these female participants increasing the count of shootings around New York through the analysis.
